---
title: "analysis-script"
author: "Ganesh Shelke"
date: "15/06/2019"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

## Libraries

```{r}
library(tidyverse)
library(lubridate)
library(dslabs)
library(caret)
```

#### Data Import

We have used MovieLens 10M movie ratings. The MovieLens data set contains 10000054 rows, 10677 movies, 797 genres and 69878 users. Released 1/2009. The data were imported from GroupLens using a script provided by the course. After removing 10% validation set, the 90% training set will be referred to as the edX dataset to disambiguate it from training and test sets derived from this EdX dataset. The version of movielens included in the dslabs package is just a small subset of a much larger dataset with millions of ratings.


```{r}

###################################
# Create edx set and validation set
###################################

# Note: this process could take a couple of minutes

if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")

# MovieLens 10M dataset:
# https://grouplens.org/datasets/movielens/10m/
# http://files.grouplens.org/datasets/movielens/ml-10m.zip

dl <- tempfile()
download.file("http://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)

ratings <- read.table(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
                      col.names = c("userId", "movieId", "rating", "timestamp"))

movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), "\\::", 3)
colnames(movies) <- c("movieId", "title", "genres")
movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(levels(movieId))[movieId],
                                           title = as.character(title),
                                           genres = as.character(genres))

movielens <- left_join(ratings, movies, by = "movieId")

# Validation set will be 10% of MovieLens data

set.seed(1, sample.kind = "Rounding")
test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
edx <- movielens[-test_index,]
temp <- movielens[test_index,]

# Make sure userId and movieId in validation set are also in edx set

validation <- temp %>% 
  semi_join(edx, by = "movieId") %>%
  semi_join(edx, by = "userId")

# Add rows removed from validation set back into edx set

removed <- anti_join(temp, validation)
edx <- rbind(edx, removed)

rm(dl, ratings, movies, test_index, temp, movielens, removed)

```


### Data Wrangling

We have performed some wrangling steps on both the edX and validation sets before splitting it into training and test data and the following variables were generated (with its type in bracket)

* `action` - whether film includes "Action" genre (logical)
* `adventure` - whether film includes "Adventure" genre (logical)
* `animation` - whether film includes "Animation" genre (logical)
* `children` - whether film includes "Children" genre (logical)
* `comedy` - whether film includes "Comedy" genre (logical)
* `crime` - whether film includes "Crime" genre (logical)
* `documentary` - whether film includes "Documentary" genre (logical)
* `drama` - whether film includes "Drama" genre (logical)
* `fantasy` - whether film includes "Fantasy" genre (logical)
* `filmNoir` - whether film includes "Film-Noir" genre (logical)
* `horror` - whether film includes "Horror" genre (logical)
* `imax` - whether film includes "IMAX" genre (logical)
* `musical` - whether film includes "Musical" genre (logical)
* `mystery` - whether film includes "Mystery" genre (logical)
* `romance` - whether film includes "Romance" genre (logical)
* `sciFi` - whether film includes "Sci-Fi" genre (logical)
* `thriller` - whether film includes "Thriller" genre (logical)
* `war` - whether film includes "War" genre (logical)
* `western` - whether film includes "Western" genre (logical)
* `unknown` - whether film includes "Unknown" genre (logical)
* `ratingFactor` - star rating as a factor (factor)

### edx dataset : it's 90% of the 10M Movielens dataset.

#### Adding `ratingFactor`

```{r}
edx <- edx %>%
  mutate(ratingFactor = factor(rating))
```


#### Adding base genre variables to edx

```{r}
edx <- edx %>%
  mutate(action = str_detect(genres, "Action"),
         adventure = str_detect(genres, "Adventure"),
         animation = str_detect(genres, "Animation"),
         children = str_detect(genres, "Children"),
         comedy = str_detect(genres, "Comedy"),
         crime = str_detect(genres, "Crime"),
         documentary = str_detect(genres, "Documentary"),
         drama = str_detect(genres, "Drama"),
         fantasy = str_detect(genres, "Fantasy"),
         filmNoir = str_detect(genres, "Film-Noir"),
         imax = str_detect(genres, "IMAX"),
         horror = str_detect(genres, "Horror"),
         musical = str_detect(genres, "Musical"),
         mystery = str_detect(genres, "Mystery"),
         romance = str_detect(genres, "Romance"),
         sciFi = str_detect(genres, "Sci-Fi"),
         thriller = str_detect(genres, "Thriller"),
         war = str_detect(genres, "War"),
         western = str_detect(genres, "Western"),
         unknown = str_detect(genres, "Unknown"))
```

####`userId` and `movieId` are converted to factors.

```{r}
edx <- edx %>%
  mutate(userId = factor(userId),
         movieId = factor(movieId))
```

#Now we will perfom the same steps for validation data set which is 10% (999,999 ratings) of the 10M Movielens dataset.

### validation dataset wrangling

#### Adding `ratingFactor`

```{r}
validation <- validation %>%
  mutate(ratingFactor = factor(rating))
```

#### Adding base genre variables to validation data

```{r}
validation <- validation %>%
  mutate(action = str_detect(genres, "Action"),
         adventure = str_detect(genres, "Adventure"),
         animation = str_detect(genres, "Animation"),
         children = str_detect(genres, "Children"),
         comedy = str_detect(genres, "Comedy"),
         crime = str_detect(genres, "Crime"),
         documentary = str_detect(genres, "Documentary"),
         drama = str_detect(genres, "Drama"),
         fantasy = str_detect(genres, "Fantasy"),
         filmNoir = str_detect(genres, "Film-Noir"),
         imax = str_detect(genres, "IMAX"),
         horror = str_detect(genres, "Horror"),
         musical = str_detect(genres, "Musical"),
         mystery = str_detect(genres, "Mystery"),
         romance = str_detect(genres, "Romance"),
         sciFi = str_detect(genres, "Sci-Fi"),
         thriller = str_detect(genres, "Thriller"),
         war = str_detect(genres, "War"),
         western = str_detect(genres, "Western"),
         unknown = str_detect(genres, "Unknown"))
```

#### `userId` and `movieId` are converted to factors and checked for levels of those factors with edx data

```{r}
validation <- validation %>%
  mutate(userId = factor(userId , levels = levels(edx$userId)),
         movieId = factor(movieId, levels = levels(edx$movieId)))
```

## Training and test sets

The EdX 10M MovieLens data were split into a 90% training set (edx dataset) and 10% test set (validation data). The test set contained only movies and users also present in the training set.


## Modeling 

### The Naive assumption - to predict the mean rating

```{r}
# calculate muHat
muHat <- mean(edx$rating)
# calculating RMSE for naive model of predicting muHat
naiveRmse <- RMSE(edx$rating, muHat)
# defining results table (we will add results of different methods to this table)
resultsTable <- data_frame(method = "Mean rating", RMSE = naiveRmse)
```

### Movie effect with regularization

```{r}
# The 5-folds cross-validation
set.seed(1, sample.kind = "Rounding")
crossValFolds <- createFolds(edx$rating, k = 5)
# calculating regularization coefficient for movie effect on 5 folds and storing it as movieLambdas
movieLambdas <- map_dbl(1:5, function(i){
  # creating foldTrain and foldTest, local train and test sets for the ith fold
  index <- crossValFolds[[i]]
  foldTrain <- edx[index, ]
  foldTest <- edx[-index, ]
  
  # remove movies from foldTest that are not present in foldTrain
  foldTest <- foldTest %>%
    filter(movieId %in% foldTrain$movieId)
  
  # assign lambdas to try
  lambdas <- seq(0, 10, 0.25)
  
  # define sumResiduals and numReviews for regularization formula
  movieResidualInfo <- foldTrain %>%
    group_by(movieId) %>%
    summarize(sumResiduals = sum(rating - muHat), numReviews = n())
  
  # try a variety of lambdas as regularization coefficients and calculate RMSE for each lambda
  rmses <- sapply(lambdas, function(lambda){
    predicted_ratings <- foldTest %>%
      left_join(movieResidualInfo, by = "movieId") %>%
      mutate(movieBias = sumResiduals/(numReviews + lambda),
             pred = muHat + movieBias) %>%
      .$pred
    return(RMSE(predicted_ratings, foldTest$rating))
  })
  
  # return the lambda that generates the minimum RMSE
  lambdas[which.min(rmses)]
})
# Taking the mean of the movie regularization coefficients across folds and store as movieLambda
movieLambda <- mean(movieLambdas)
# calculate regularized movie effect for each movie using optimized regularization coefficient, store as movieBiasRegularized
movieBiasRegularized <- edx %>%
  group_by(movieId) %>%
  summarize(movieBiasReg = sum(rating - muHat)/(n() + movieLambda))
```

```{r}
# join movie bias to training set
edx <- edx %>%
  left_join(movieBiasRegularized)
# join movie bias to test set
validation <- validation %>%
  left_join(movieBiasRegularized)
# calculate test set ratings with muHat and movie bias
movieEffectRegModel <- muHat + validation$movieBiasReg
# calculate test set RMSE
movieEffectRegRMSE <- RMSE(validation$rating, movieEffectRegModel)
# add movie effect RMSE to results table
resultsTable <- bind_rows(resultsTable,
                          data_frame(method = "Regularized movie effect added",
                                     RMSE = movieEffectRegRMSE))
```


# Now we will add other effects to the regularized movie effect we have already calculated: regularized user effect and regularized genre effect and finally apply this model to our test set that is validation set to report final RMSE on validation set.

### User effect with regularization

```{r}
## define regularized user effects
# calculate regularization coefficient for user effect on 5 folds, store as userLambdas
userLambdas <- map_dbl(1:5, function(i){
  # create foldTrain and foldTest, local train and test sets for the ith fold
  index <- crossValFolds[[i]]
  foldTrain <- edx[index, ]
  foldTest <- edx[-index, ]
  
  # remove users from foldTest that are not present in foldTrain
  foldTest <- foldTest %>%
    filter(userId %in% foldTrain$userId)
  
  # assign lambdas to try
  lambdas <- seq(0, 10, 0.25)
  
  # define sumResiduals and numReviews for regularization formula
  userResidualInfo <- foldTrain %>%
    group_by(userId) %>%
    summarize(sumResiduals = sum(rating - muHat - movieBiasReg), numReviews = n())
  
  # try a variety of lambdas as regularization coefficients and calculate RMSE for each lambda
  rmses <- sapply(lambdas, function(lambda){
    predicted_ratings <- foldTest %>%
      left_join(userResidualInfo, by = "userId") %>%
      mutate(userBias = sumResiduals/(numReviews + lambda),
             pred = muHat + movieBiasReg + userBias) %>%
      .$pred
    return(RMSE(predicted_ratings, foldTest$rating))
  })
  
  # return the lambda that generates the minimum RMSE
  lambdas[which.min(rmses)]
})
# determine mean lambda across folds
userLambda <- mean(userLambdas)
# calculate regularized user effect for each movie using optimized regularization coefficient, store as userBiasRegularized
userBiasRegularized <- edx %>%
  group_by(userId) %>%
  summarize(userBiasReg = sum(rating - muHat - movieBiasReg)/(n() + userLambda))
```

```{r}
# join user bias to train set
edx <- edx %>%
  left_join(userBiasRegularized)
# join user bias to test set
validation <- validation %>%
  left_join(userBiasRegularized)
# calculate predictions with muHat, movie and user effects
userEffectRegModel <- muHat + validation$movieBiasReg + validation$userBiasReg
# calculate RMSE for muHat, movie and user effects
userEffectRegRMSE <- RMSE(validation$rating, userEffectRegModel)
# add user model to results table
resultsTable <- bind_rows(resultsTable,
                          data_frame(method = "Regularized user effect added",
                                     RMSE = userEffectRegRMSE))
```

### Genre effects with regularization

```{r}
# calculate regularization coefficient for user effect on 5 folds, store as userLambdas
genreLambdas <- map_dbl(1:5, function(i){
  # create foldTrain and foldTest, local train and test sets for the ith fold
  index <- crossValFolds[[i]]
  foldTrain <- edx[index, ]
  foldTest <- edx[-index, ]
  
  # remove users from foldTest that are not present in foldTrain
  foldTest <- foldTest %>%
    filter(genres %in% foldTrain$genres)
  
  # assign lambdas to try
  lambdas <- seq(100, 1000, 100)
  
  # define sumResiduals and numReviews for regularization formula
  genreResidualInfo <- foldTrain %>%
    group_by(genres) %>%
    summarize(sumResiduals = sum(rating - muHat - movieBiasReg - userBiasReg), numReviews = n())
  
  # try a variety of lambdas as regularization coefficients and calculate RMSE for each lambda
  rmses <- sapply(lambdas, function(lambda){
    predicted_ratings <- foldTest %>%
      left_join(genreResidualInfo, by = "genres") %>%
      mutate(genreBias = sumResiduals/(numReviews + lambda),
             pred = muHat + movieBiasReg + userBiasReg + genreBias) %>%
      .$pred
    return(RMSE(predicted_ratings, foldTest$rating))
  })
  
  # return the lambda that generates the minimum RMSE
  lambdas[which.min(rmses)]
})
# determine mean lambda across folds
genreLambda <- mean(genreLambdas)
# calculate regularized user effect for each movie using optimized regularization coefficient, store as userBiasRegularized
genreBiasRegularized <- edx %>%
  group_by(genres) %>%
  summarize(genreBiasReg = sum(rating - muHat - movieBiasReg - userBiasReg)/(n() + userLambda))
```


```{r}
# join genre effects to train set
edx <- edx %>%
  left_join(genreBiasRegularized)
# join genre effects to test set
validation <- validation %>%
  left_join(genreBiasRegularized)
# calculate test predictions with muHat, movie bias, user bias and genre bias
genreEffectRegModel <- muHat + validation$movieBiasReg + validation$userBiasReg + validation$genreBiasReg
# calculate test RMSE with muHat, movie bias, user bias and genre bias
genreEffectRegRMSE <- RMSE(validation$rating, genreEffectRegModel)
# add genre effect to results table
resultsTable <- bind_rows(resultsTable,
                          data_frame(method = "Regularized genre effect added",
                                     RMSE = genreEffectRegRMSE))
```

### Applying final model to validation set

```{r}
# join movie, user and genre bias to validation set
validation <- validation %>%
  left_join(movieBiasRegularized) %>%
  left_join(userBiasRegularized) %>%
  left_join(genreBiasRegularized)
# calculate predictions for validation set on final model
finalModel <- muHat + validation$movieBiasReg + validation$userBiasReg + validation$genreBiasReg
# calculate RMSE on final model
finalRMSE <- RMSE(validation$rating, finalModel)
# add genre effect to results table
resultsTable <- bind_rows(resultsTable,
                          data_frame(method = "Final model on validation set",
                                     RMSE = finalRMSE))
```

## The final RMSE

```{r}
finalRMSE
```
